---
title: \vspace{5cm} Predicting Vaccine Hesitancy for Humana Members
author: 2021 Humana-Mays Healthcare Analytics Case Competition
date: October 10, 2021
output: 
  pdf_document: 
    number_sections: yes
    df_print: kable
    keep_tex: yes
abstract: 
header-includes:
 \usepackage{setspace}
 \doublespacing
 \setlength{\parskip}{2mm}
 \usepackage{float}
 \floatplacement{figure}{H}
 \usepackage[skip=4pt]{caption}
bibliography: ref.bib
csl: apa.csl
indent: true
geometry: margin=1.25in
fontsize: 12pt
---

\newpage
\tableofcontents

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      echo = FALSE,
                      warning = FALSE,
                      fig.pos = "H",
                      fig.align = "center")
```

```{r}
library(tidyverse)
library(tidymodels)
theme_set(theme_light())
```

\newpage

# Executive Summary

This study was a statistical data investigation into COVID-19 vaccination status for Humana members. In particular, our main goal was to build a predictive model for identifying which members are most likely to be hesitant to the COVID-19 vaccine. Furthermore, we wanted to provide recommendations and potential solutions to increase vaccination rates among the sub-segments of hesitant members based on the insights obtained from the data.

In this work, we implemented two boosted tree classification methods, namely, XGBoost and CatBoost, to predict vaccination status for Humana members and identify important variables that contribute the most to the predictions. After preprocessing the data and training the models, we found that CatBoost was the best classifier in terms of maximizing the area under a receiver operating characteristic (ROC) curve (AUC). In addition, we obtained from our best model that Member Age and Race were the two most significant predictors for explaining the likelihood of vaccine hesitancy.

Then, based on our observations, we recommended that Humana distributes pamphlets and surveys, utilizes peer-to-peer communication, and performs joint-intervention with pharmacies to target those refusing the vaccine. By implementing our recommendations based on the model predictions, key indicators, and relevant research, Humana will be able to achieve its goal of increasing the COVID-19 vaccination rate among their members.

\newpage

# Background

Humana is a leading healthcare company that offers a wide range of insurance products and health and wellness services. Their mission is simply to help people achieve lifelong well-being. In addition, with the current COVID-19 pandemic, increasing the vaccination rates among the members of Humana remains to be the top priority for both members as well as the larger population. They are primarily focused on providing vaccination opportunities for the most vulnerable and underserved populations. Thus, the objective of this study is to help Humana identify which members are likely to be hesitant so that they could design targeted outreaches for these individuals.

Extending into nearly every field from sports, education, to general healthcare, vaccine hesitancy is an issue that has crept into the minds of people beyond the business sector.  Nearly 80% of the unvaccinated are unlikely to change their minds about getting a vaccine [@gallup], highlighting the importance of addressing vaccine confidence today. Our goal is to build a predictive model using observed characteristics provided by the data to estimate the probability of being hesitant to receive the COVID-19 vaccine for each member. More importantly, with the use of these probabilities, features, and statistical analysis, we wanted to provide recommendations and potential solutions to drive vaccination among the sub-segments of hesitant members. 

\newpage

# Data

The training dataset provided for this study contains a total of 974,842 rows and 368 columns in which information on medical claims, pharmacy claims, lab claims, demographic/consumer data, credit data, condition-related features, CMS features, and other features were included. Our target variable, `covid_vaccination` indicates the vaccination status of each member. Vaccinated members were coded as `vacc` and  unvaccinated members were coded as `no_vacc`.

We first wanted to reduce the size of our data, in particular, the number of columns. For instance, we filtered out variables with (near)-zero variance to eliminate potential problems with model stability. In addition, we repaired categorical variables that were misrepresented as numeric features when the data was imported. This left us with 311 total independent variables and 1 column for the response variable - whether or not a member is vaccinated.

After cleaning up the data, we summarized the COVID vaccination status for members and looked for trends among various predictors. First, we looked at the relationship between whether or not a member is vaccinated versus age and gender. We decided to convert age, which was originally a continuous variable, into discrete intervals (bins), for the purpose of exploratory analysis only. The age groups that we partitioned age into are 20-55, 55-65, 65-75, and 75+. We could see from Figure 1 that older members tend to have higher vaccination rate than younger members. Moreover, within each age group, the vaccination percentage is higher for female members than the male group.

```{r}
# see cb/eda files for splits
splits <- read_rds("splits.rds")
train <- training(splits)
test <- testing(splits)
```

```{r, fig.cap="Bar graph of percent vaccinated by age group, color coded by gender.", fig.height=3.6, fig.width=4.8}
train %>% 
  select(covid_vaccination, est_age, sex_cd) %>% 
  mutate(age_group = cut(est_age, 
                         include.lowest = TRUE, 
                         breaks = c(20, 55, 65, 75, 104),
                         labels = c("[20-55]", "(55-65]", "(65-75]", "75+"))) %>% 
  group_by(age_group, sex_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  ggplot(aes(pct_vac, age_group, fill = sex_cd)) + 
  geom_col(position = "dodge") +
  scale_x_continuous(labels = scales::percent) +
    scale_fill_manual(values = c("#5BA908", "#003182")) +
  labs(y = "Age Group",
       x = "Percent Vaccinated",
       fill = "Gender")
```

Another variable we wanted to explore is race. More than 80% of the members in the original data were identified as White, with the rest being Black, Other, Asian, Hispanic, North American Native, or Unknown. Figure 2 tells us that members with race codes 5, 6, and 2, which are Hispanic, North American Native, and Black, respectively, have lower vaccination rate than other race groups.

In addition, we looked at the association between vaccination status and the original reason for entry into medicare [@medicare]. According to Figure 3, members who entered Medicare because of Old Age Survivors Insurance (OASI, coded 0) obtained the highest vaccination rate out of all the other reasons (Disability Insurance Benefits (DIB), End-stage Renal Disease (ESRD), or both - coded 1, 2, and 3).

```{r, fig.cap="Bar graph of percent vaccinated by race. The race code values are: 0 = Unknown; 1 = White; 2 = Black; 3 = Other; 4 = Asian; 5 = Hispanic; 6 = North American Native", fig.height=3.2, fig.width=4}
train %>% 
  select(covid_vaccination, race_cd) %>%
  drop_na() %>% 
  group_by(race_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  mutate(race_cd = as.factor(race_cd),
         race_cd = fct_reorder(race_cd, pct_vac)) %>% 
  ggplot(aes(pct_vac, race_cd)) + 
  geom_col(position = "dodge", show.legend = FALSE, fill = "#5BA908") +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Race",
       x = "Percent Vaccinated")
```

```{r, fig.cap = "Bar graph of percent vaccinated by original reason for entry into Medicare. The original reason for Medicare entitlement code values are: 0	= Old age and survivor’s insurance (OASI); 1 = Disability insurance benefits (DIB); 2 = End-stage renal disease (ESRD); 3	= Both DIB and ESRD", fig.height=3.2, fig.width=4}
train %>% 
  select(covid_vaccination, cms_orig_reas_entitle_cd) %>% 
  drop_na() %>% 
  group_by(cms_orig_reas_entitle_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  mutate(cms_orig_reas_entitle_cd = as.factor(cms_orig_reas_entitle_cd),
         cms_orig_reas_entitle_cd = fct_reorder(cms_orig_reas_entitle_cd, pct_vac)) %>% 
  ggplot(aes(pct_vac, cms_orig_reas_entitle_cd)) + 
  geom_col(position = "dodge", show.legend = FALSE, fill = "#5BA908") +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Medicare entry reason",
       x = "Percent Vaccinated")
```

\newpage

# Predictive Models

In this section, we presented our models for predicting vaccine hesitancy for Humana members. To tackle this prediction problem, we considered the following supervised machine learning techniques: XGBoost (eXtreme Gradient Boosting) and CatBoost. To evaluate the performance of the classfiers, the metric that we used was area under the receiver operating characteristic curve (ROC/AUC).

## XGBoost

The first model we considered was gradient boosted trees using the popular eXtreme Gradient Boosting (XGBoost) method [@xgpaper]. We implemented XGBoost via the `xgboost` package [@xgpackage] and with the `tidymodels` framework [@tidymodels] in the `R` programming language [@rcite].  To get the data into a useful format, median imputation was performed on numeric variables with missing data, while an "unknown" level representing missing values was added to the categorical covariates. We also implemented one-hot encoding to turn our qualitative predictors into indicator columns.

After that, we trained our XGBoost model on a 75-25 train-test split, with 3-fold cross-validation. Our final XGBoost model consisted of the following hyperparameters: learning rate of 0.02, 1150 trees, and a number of predictors sampled for spliting at each node of 6. We then made predictions on the validation set and obtained an ROC/AUC score of 0.6727 from this XGBoost fit (see Figure 4).

```{r, fig.cap="Receiver operating characteristic curve for XGBoost classification performance on the validation set.", fig.height = 4, fig.width = 4}
read_csv("xg_roc.csv") %>% 
  ggplot(aes(1 - specificity, sensitivity)) +
  geom_abline(slope = 1, lty = 2) +
  geom_path(size = 1.5, color = "#5BA908") +
  labs(x = "False Positive Rate", 
       y = "True Positive Rate",
       subtitle = "AUC = 0.6727")
```


## CatBoost

In addition, we explored another gradient boosting framework, CatBoost [@cbpaper], provided by the `catboost` `R` package. The advantage of using CatBoost is that minimal data preparation is required. This algorithm handles missing values for numeric features and non-encoded categorical features. The only data mutating steps we performed were turning the response (`covid_vaccination`) into a binary indicator with 0 = `vacc` and 1 = `no_vacc` (since the objective is to identify vaccine hesitant members), and converting categorical features into "factors" (a type of data object for categorical variables in `R`).

For our CatBoost model, we first partitioned the data into a training set and a validation set, using a 75-25 split ratio, similar to what we did for XGBoost. We found that the combination of learning rate of 0.05 and the default settings for the rest of the parameters yielded the best results. We utilized our best CatBoost model to get the predicted probability values for vaccine hesitancy on the validation set, and achieved an ROC/AUC value of 0.6846 (see Figure 5). Compared to the previous fit, the CatBoost model outperformed XGBoost, in terms of ROC/AUC. 

```{r, fig.cap="Receiver operating characteristic curve for CatBoost classification performance on the validation set.", fig.height = 4, fig.width = 4}
read_csv("cb_test_roc.csv") %>%
  roc_curve(as.factor(covid_vaccination), pred) %>%
  ggplot(aes(1 - specificity, sensitivity)) +
  geom_abline(slope = 1, lty = 2) +
  geom_path(size = 1.5, color = "#5BA908") +
  labs(x = "False Positive Rate", 
       y = "True Positive Rate",
       subtitle = "AUC = 0.6846")
```


## Feature Importance

After fitting the classification models, we wanted to look at how much the models rely on the features to make predictions. Since CatBoost gave us better classification results, we decided to look at the variable importance obtained from this model (see Figure 6). 

```{r, fig.cap="Top 20 important features from CatBoost model. The importance of each variable was calculated based on how much on average the prediction changes if the feature value changes.", fig.with=7}
imp <- read_csv("var_imp_cb_final.csv")
imp %>% 
  slice_max(n = 20, order_by = imp) %>%
  mutate(var = fct_reorder(var, imp)) %>%
  ggplot(aes(x = var, y = imp)) +
  geom_col(fill = "#5BA908") +
  coord_flip() +
  theme_light() +
  labs(y = "Importance",
       x = "Variable")
```

Table 1 gives full information about the top 20 important features from our CatBoost model. It is clear that age is the variable that contributes the most to the predictions of vaccination hesitancy. As we have shown earlier in Section \ref{data}, younger member tend to have lower vaccination rate than elderly people. In addition, race is also a significant factor in identifying whether a member is vaccinated. We discovered earlier that Hispanic, North American Native, and Black are the race groups that have low vaccination rate. 

\singlespacing

\small

```{r}
library(kableExtra)

info <- c("Member age (calculated using est_bday, relative to score/index date)",
  "Code indicating a member's race (0 = Unknown, 1 = White, 2 = Black, 3 = Other, 4 = Asian, 5 = Hispanic, 6 = N. American Native)",
  "Code indicating the original reason for entry into Medicare",
  "Household very low food security (%, three-year average), 2013-15",
  "Risk Adjustment Factor A Amount",
  "Child & Adult Care (% pop)",
  "Household food insecurity (%, three-year average), 2013-15",
  "WIC participants (% pop)",
  "RX Adherence - Maintenance",
  "Trend of cost per month of prescriptions related to vaccine drugs in the past sixth to ninth month versus ninth to twelfth month prior to the score date (Based on GPI2 grouping)",
  "Expenditures per capita, fast food",
  "Summer Food Service Program participants (% pop)",
  "Estimated Household Investable Assets Recoded",
  "Member geographic information - Humana Region",
  "National School Lunch Program participants (% pop)",
  "Days since last prescription in the past three
months prior to score date",
  "Net Worth Per Adult",
  "Clinical Care - Percentage of children under age 19
without health insurance",
  "Count per month of prescriptions related to
maintenance drugs in the past ninth to twelfth
month prior to the score date",
  "SNAP participants (% pop)"
)

imp %>% 
  slice_max(n = 20, order_by = imp) %>% 
  mutate(var = ifelse(var == "cms_orig_reas_entitle_cd",
                      "cms_orig_reas _entitle_cd", var),
         var = ifelse(var == "cms_risk_adjustment_factor_a_amt",
                      "cms_risk_adjustment _factor_a_amt", var),
         var = ifelse(var == "rx_gpi2_17_pmpm_cost_t_12_9_6m_b4",
                      "rx_gpi2_17_pmpm_cost _t_12_9_6m_b4", var),
         var = ifelse(var == "rx_days_since_last_script_0to3m_b4", "rx_days_since_last _script_0to3m_b4", var),
         var = ifelse(var == "rx_maint_pmpm_ct_9to12m_b4", "rx_maint_pmpm_ct _9to12m_b4", var)
         ) %>% 
  transmute(Feature = var,
            Description = info) %>% 
  kable(caption = "Complete descriptions of the top 20 important features obtained from CatBoost fit. These information were extracted from the data dictionary provided by the competition.",
        booktabs = TRUE,
        position = "H",
        format = "latex") %>% 
  column_spec(1, width = "2in") %>% 
  column_spec(2, width = "3.6in")
```

\normalsize

\doublespacing

\newpage

# Recommendations

According to a recent Gallup poll of 3,572 Americans [@gallup], 78% (of the unvaccinated surveyed) indicated that they are unlikely to or definitely will not change their minds in receiving the COVID-19 vaccine. This provides an immense challenge to healthcare providers as far as recommendations go, raising the question of how they can convince non-vaccinated individuals to get the vaccine when they are already adamant about their stance. Building a model using modern classification techniques to predict vaccination status is only getting Humana halfway there. 

A crucial finding from Gallup’s poll was that, of the unvaccinated surveyed, 23% cited the main reason for not getting vaccinated was they were concerned about the safety of the vaccine. While 78% are unlikely to change their minds, perhaps an approach that administers positive and accurate information and improves knowledge dissemination will allow us to improve vaccination rate. Before considering the results of our analysis, we would recommend Humana to mail out pamphlets related to vaccine safety to those in their service regions to ease the main concerns of those refusing vaccination.

Moreover, we could target these mailings to the demographic that is currently likely to be unvaccinated. Younger people should be considered a high priority for this initiative (see Figure 1). One possible reason that younger people were found to be less likely to receive the vaccine is that they do not face considerable risk to their survival and overall well-being compared to infantile and elderly populations. Under this hypothesis, we would also recommend that Humana identify unvaccinated young adults and provide them with a realistic overview of the effects of COVID-19 beyond individual harm. Instead, we should highlight the real issues of economic pitfalls, intensive care units at capacity, and a lack of outbreak detection in hospices that treat the elderly [@impact], all of which could be avoided through vaccination. 

Additionally, an essential target found in our model includes racial and ethnic minority groups (Black, Hispanic, and Native Americans), since these groups appeared to have relatively lower vaccination rates compared to White and Non-Hispanic groups (see Figure 2). According to The Centers for Disease Control and Prevention (CDC), this is due to challenges including socioeconomic status, gaps in healthcare access, transportation issues, lack of trust because of past medical racism and experimentation, and other factors [@cdc]. In order to tackle these challenges, that is, emphasize vaccine equity for racial and ethnic minority groups, we suggest that Humana build rapport among their members in these groups before even initiating to educate them. Through this, we may have higher chances of them listening and gaining their trust and confidence in the vaccine.

At the same time, we must empathize with those that choose not to have themselves or their children vaccinated. Therefore, Humana should make these messages without using phrases such as "should get the vaccine." Instead, we should use diction that depicts Humana’s understanding of the concerns among unvaccinated individuals. According to a World Economic Forum report on building trust in vaccines [@trust], messages that make vaccines seem as if they are a moral obligation will often evoke negative emotions. Successful messages will not rely on morality-based arguments involving celebrities and politicians, but rather arguments based on gratitude from healthcare professionals or "people like me." To maximize the effectiveness of our pamphlets, Humana should incorporate these insights while making messages regarding the protection of the elderly, the economy, the healthcare system, and of course, those receiving the vaccine.

Another high priority target group includes members who obtain prescriptions related to maintenance drugs in our model to predict COVID-19 vaccination status, as these individuals contribute to a considerable amount of our unvaccinated group. Thus, we recommend that Humana establish a joint intervention with pharmacies, including pamphlets supporting COVID-19 vaccination in members’ prescription drug packages. In doing so, we hope to target a large subpopulation that has relatively low vaccination rates.  

In an even greater effort to encourage the elderly to receive the COVID-19 vaccine, we suggest that Humana targets areas such as community centers or nursery homes in unvaccinated “hotspot” areas. Informative presentations about the benefits of the vaccine, with testimonials about COVID vaccination from their peers, could change the minds of those refusing the vaccine, given that peer-to-peer interventions have been shown to be effective based on a recent research from @peer. In order to further strengthen this approach, Humana could offer small incentives to those in attendance of the event, as well as reward those who recruit others to participate. In addition to these efforts, we recommend pre and post-surveys about COVID-19 vaccination, as a measure of these presentations. 

Along with providing this information, Humana should also send out surveys to hone in on other reasons why people are choosing to be unvaccinated - especially young men and those in the Black, Hispanic, and Native American communities. These surveys could also assist in reaching more unvaccinated Americans than we anticipate. Similar to how the spouses of smokers are more likely to be smokers themselves [@smokers], we imagine that family households with one unvaccinated family member will likely have other unvaccinated family members in their home. Lastly, it is critical to build trust between communities and healthcare/insurance providers. In this survey process, it would be wise to observe whether a lack of communication is a factor in whether or not a person is vaccinated.  After determining other various reasons for vaccine hesitancy, we can continue to update our information to address these concerns and further educate the public on the importance of vaccination. 


\newpage

# Conclusion

Humana's mission is to help people achieve lifelong well-being. COVID-19 has introduced the challenge in increasing both vaccination rates and vaccine confidence at the forefront of minds around the world. Humana is eager to achieve an increase in vaccination rates and vaccine confidence among their members (especially the younger people) as well as the larger population. Thus, for this year’s case study, we developed machine learning models that identify which members are more likely to be hesitant to the COVID-19 vaccine. 

With the provided dataset consisting of  974,842 rows and 368 columns, we fitted gradient boosted tree models to predict vaccine hesitancy for Humana members. Specifically, we considered two algorithms, XGBoost and CatBoost, with CatBoost ended up being the classifier that provided the best prediction performance. Moreover, we determined from our CatBoost model that age and race were the two variables that have the greatest contributions on predicting whether or not a member is vaccinated.

Based on the important features obtained from our model, we provide the following recommendations to build confidence in the COVID-19 vaccine and increase vaccination rates: 

1.    Mail out pamphlets that summarize accurate and positive information related to the safety and the benefits of the vaccine.

2.    Focus on unvaccinated young adults by providing them a realistic overview of the effects of COVID-19 beyond individual harm.

3.    Build rapport among their members in the racial and ethnic minority groups before even initiating to educate them.

4.    Establish a joint intervention with pharmacies by including pamphlets supporting COVID-19 vaccination in members’ prescription drug packages.

5.    Target community centers or nursery homes in unvaccinated "hotspot" areas and provide an effective and informative presentation regarding the benefits of the vaccine.

6.    Administer pre and post-surveys about COVID-19 vaccination, as a measure of effectiveness of these presentations.

7.    Conduct surveys to obtain more information and understanding on COVID-19 hesitancy.

It is imperative for healthcare and insurance providers to build a strong relationship with the communities before even starting to inform and educate them. With these fundamental relationships established, only then will the public be able to invest its willingness and trust in our healthcare agencies. And along with their trust comes the ability for the community to achieve Humana’s mission of lifelong well-being. 

\newpage

# References

