---
title: "EDA Figures"
output: 
  html_document: 
    df_print: kable
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      comment = "")
```

## Set up

### Packages

```{r}
library(tidyverse)
library(tidymodels)
library(janitor)
theme_set(theme_light())
```

### Data

```{r, eval = FALSE}
train_raw <- read_csv("data/2021_Competition_Training.csv", 
                      na = c("*", "")) %>% 
  select(-1) %>% 
  clean_names()

test_raw <- read_csv("data/2021_Competition_Holdout.csv", 
                     na = c("*", "")) %>% 
  select(-1) %>% 
  clean_names()
```

### Splits before EDA 

(EDA only on the train set)

```{r, eval = FALSE}
set.seed(100)
splits <- train_raw %>% 
  initial_split(strata = covid_vaccination)

train <- training(splits)
test <- testing(splits)
```

```{r, echo = FALSE}
splits <- read_rds("splits.rds")
```

```{r, echo = FALSE}
train <- training(splits)
test <- testing(splits)
```


## EDA

### Age group + Gender

```{r}
range(train$est_age)
```

```{r}
train %>% 
  select(covid_vaccination, est_age, sex_cd) %>% 
  mutate(age_group = cut(est_age, 
                         include.lowest = TRUE, 
                         breaks = c(20, 55, 65, 75, 104),
                         labels = c("[20-55]", "(55-65]", "(65-75]", "75+"))) %>% 
  group_by(age_group, sex_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  ggplot(aes(pct_vac, age_group, fill = sex_cd)) + 
  geom_col(position = "dodge") +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Age Group",
       x = "Percent Vaccinated",
       fill = "Gender")
```

Clearly a trend here, older people tend to have higher vaccination rate.

Within each age group, women also have higher vaccination rate than men. This trend also holds overall (no Simpson paradox here)

```{r}
train %>% 
  select(covid_vaccination, sex_cd) %>% 
  group_by(sex_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n)
```

### `src_div_id`: The division ID assigned by the legacy system {each
platform has a different system}

```{r}
train %>% 
  select(covid_vaccination, src_div_id) %>% 
  mutate(src_div_id = str_remove(src_div_id, "00")) %>% 
  drop_na() %>% 
  group_by(src_div_id) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  mutate(src_div_id = fct_reorder(src_div_id, pct_vac)) %>% 
  ggplot(aes(pct_vac, src_div_id, fill = src_div_id)) + 
  geom_col(position = "dodge", show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Division ID ",
       x = "Percent Vaccinated")
```

Divisions 0 and 1 have the lowest vaccinate rate.

### `cms_orig_reas_entitle_cd`: Code indicating the original reason for entry into Medicare

```{r}
train %>% 
  select(covid_vaccination, cms_orig_reas_entitle_cd) %>% 
  drop_na() %>% 
  group_by(cms_orig_reas_entitle_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  mutate(cms_orig_reas_entitle_cd = as.factor(cms_orig_reas_entitle_cd),
         cms_orig_reas_entitle_cd = fct_reorder(cms_orig_reas_entitle_cd, pct_vac)) %>% 
  ggplot(aes(pct_vac, cms_orig_reas_entitle_cd, fill = cms_orig_reas_entitle_cd)) + 
  geom_col(position = "dodge", show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Original reason for Medicare entry code",
       x = "Percent Vaccinated")
```

Code 0 have higher rate than the rest

### `cms_risk_adjustment_factor_a_amt`: Risk Adjustment Factor A Amount

Look at histogram

Original variable

```{r}
train %>% 
  select(covid_vaccination, cms_risk_adjustment_factor_a_amt) %>% 
  drop_na() %>% 
  ggplot(aes(cms_risk_adjustment_factor_a_amt, fill = covid_vaccination)) +
  geom_histogram(alpha = 0.8)
```

Distribution is kinda skewed, so try a log base 10 transformation

```{r}
train %>% 
  select(covid_vaccination, cms_risk_adjustment_factor_a_amt) %>% 
  drop_na() %>% 
  ggplot(aes(cms_risk_adjustment_factor_a_amt, fill = covid_vaccination)) +
  geom_histogram(alpha = 0.8) +
  scale_x_log10()
```

### `zip_cd`: Zip code

Top 25 zip code with highest vaccination %

```{r}
train %>% 
  select(covid_vaccination, zip_cd) %>% 
  mutate(zip_cd = as.character(zip_cd)) %>% 
  drop_na() %>% 
  group_by(zip_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  slice_max(n = 25, order_by = n) %>% 
  mutate(zip_cd = ifelse(str_length(zip_cd) == 3, str_c("00", zip_cd), zip_cd),
         zip_cd = fct_reorder(zip_cd, pct_vac)) %>% 
  ggplot(aes(pct_vac, zip_cd, fill = zip_cd)) + 
  geom_col(show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent) +
    labs(y = "Zip code",
       x = "Percent Vaccinated")
```

Where are these places???

```{r}
library(zipcodeR)
```

```{r}
train %>% 
  select(covid_vaccination, zip_cd) %>% 
  mutate(zip_cd = as.character(zip_cd)) %>% 
  drop_na() %>% 
  group_by(zip_cd) %>% 
  summarize(n = n(),
            n_vac = sum(covid_vaccination == "vacc"),
            pct_vac = n_vac / n) %>% 
  slice_max(n = 25, order_by = n) %>% 
  mutate(zip_cd = ifelse(str_length(zip_cd) == 3, str_c("00", zip_cd), zip_cd)) %>% 
  pull(zip_cd) %>% 
  reverse_zipcode() %>% 
  select(zipcode, post_office_city)
```

Some places couldn't be geocoded, but still...

